---
title: "Final_Project"
author: "Team 5"
date: "2024-11-21"
output:
 html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_float: yes
---

```{r include=FALSE}
library(ezids)
library(dplyr)
library(tidyverse)
library(ggplot2)
#install.packages("naniar")
library(naniar)
library(gridExtra)
library(car)
```



# EDA

## Data Cleaning

### Understand the Dataset
```{r}
# Loading the dataset
top_songs <- read.csv('universal_top_spotify_songs.csv', na.strings = c("", "NA"))

# Print out the first 5 rows of the dataset
xkabledplyhead(top_songs,title="The first 5 rows of the dataset")
```


```{r include=FALSE}
str(top_songs)
```


```{r}
colnames(top_songs)
```

```{r}
summary(top_songs)
```

1. Drop columns which are not useful for modelling and have no meaning on the popularity of a song

2. The minimum value of time signature of a song is 0, which is not possible (Remove rows with time signature = 0).

3. According to the data description, NAs in `country` means Global Top 50. Therefore, we will fill NAs with 'Global'.

4. Since the percentage of observations with NAs is relatively low at 0.046% of the observations; therefore, we will drop all the other observations with NAs.


### Drop irrelevant columns

```{r}
remove_col <- c(1) # Add on columns
top_songs <- top_songs[, -remove_col]
```


### Handling Missing Values

Visualize the missing values in each column
```{r}
gg_miss_var(top_songs)
```

Calculate the counts and percentage of missing values in each columns
```{r}
# Counts the missing values in each column
colSums(is.na(top_songs))

# Calculate the percentage of missing values in each column
col_na_percentages <-round(colSums(is.na(top_songs))/dim(top_songs)[1]*100, 3)
col_na_percentages
```


```{r}
# Impute the null values in country with 'Global'
top_songs$country = ifelse(is.na(top_songs$country), 'Global', top_songs$country)

# Drop observations with missing values
df_cleaned <- na.omit(top_songs)

# Drop observations with 0 time signature
df_cleaned <- df_cleaned[df_cleaned$time_signature!=0, ]
```

```{r}
print("Dimension of Dataset before Cleaning")
print(dim(top_songs))
print("Dimension of Dataset after Cleaning")
dim(df_cleaned)
```

### Convert Variables into Correct Data Types

```{r include=FALSE}
str(df_cleaned)
```

```{r}
# Convert columns into date data type
df_cleaned$snapshot_date <- as.Date(df_cleaned$snapshot_date, format = "%Y-%m-%d")
df_cleaned$album_release_date <- as.Date(df_cleaned$album_release_date, format = "%Y-%m-%d")

# Convert columns into factor data type
df_cleaned$time_signature <-factor(df_cleaned$time_signature, level= c(1, 3, 4, 5))
df_cleaned$is_explicit <- factor(as.logical(df_cleaned$is_explicit))
df_cleaned$mode <- factor(df_cleaned$mode)

str(df_cleaned)
```


```{r}
summary(df_cleaned)
```

### Outliers Detection

```{r}
selected_columns <- c(
  "popularity", "duration_ms", "danceability", 
  "energy", "loudness", "speechiness", 
  "acousticness", "instrumentalness", "liveness",
  "valence", "tempo")

boxplots_list <- list()

# Create box plots for each selected column
for (col in selected_columns) {
  boxplot <- df_cleaned %>% 
    ggplot(aes(y = .data[[col]])) +
    geom_boxplot(fill = "lightblue") +
    labs(title = paste("Box Plot for", col), y = col) +
    theme_minimal()
  print(boxplot)
}

```

## SMART QUESTION 1: What factors contribute most to a song's popularity over time?

### Objective: Analyze the relationship between song attributes (danceability, energy, tempo, valence, and loudness) and popularity trends over time.

```{r ,echo=TRUE,results='markup'}
spotify_data <- df_cleaned
str(spotify_data)
```


```{r ,echo=TRUE,results='markup'}

artists_list <- spotify_data %>%
  select(artists) %>%
  mutate(artists = strsplit(as.character(artists), ", ")) %>%
  unnest(artists)


artists_count <- artists_list %>%
  group_by(artists) %>%
  summarize(frequency = n()) %>%
  arrange(desc(frequency))


top_artists <- artists_count %>%
  slice_max(frequency, n = 10)


ggplot(top_artists, aes(x = reorder(artists, -frequency), y = frequency)) +
  geom_bar(stat = "identity", fill = "red") +
  theme_minimal() +
  labs(title = "Top 10 Most Common Artists Among top spotify songs",
       x = "Artist",
       y = "Frequency") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```




```{r ,echo=TRUE,results='markup'}
top_songs <- spotify_data %>%
  count(name, sort = TRUE) %>%
  slice_max(n, n = 10) %>%
  arrange(n)

# Create the bar chart
ggplot(top_songs, aes(x = reorder(name, n), y = n)) +
  geom_bar(stat = "identity", fill = "#1DB954") +  # Spotify green color
  geom_text(aes(label = n), vjust = -0.5, size = 3) +  # Add labels above bars
  labs(
    title = "Songs with Most Appearances",
    x = "",
    y = ""
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 14),
    plot.title = element_text(hjust = 0.5, size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


```


```{r ,echo=TRUE,results='markup'}
library(ggplot2)
ggplot(spotify_data, aes(x = danceability)) +
  geom_histogram(bins = 30, fill = "blue", color = "black") +
  labs(title = "Distribution of Danceability", x = "Danceability", y = "Frequency")

ggplot(spotify_data, aes(x = popularity)) +
  geom_density(fill = "purple", alpha = 0.5) +
  labs(title = "Density Plot of Popularity", x = "Popularity", y = "Density")
```

```{r ,echo=TRUE,results='markup'}
spotify_num <- spotify_data %>%
  select(popularity, danceability, energy, tempo, valence, loudness)


# Calculate correlation matrix
cor_matrix <- cor(spotify_num, use = "complete.obs")
print(cor_matrix)

# Visualize the correlation matrix
library(corrplot)
corrplot::corrplot(cor_matrix, method = "color", addCoef.col = "black")
```



The correlation between popularity and most attributes (danceability, energy, tempo, and valence) is very weak (close to 0). This suggests that the features taken alone are not strong predictors of a song's popularity.

The only attribute with a slightly positive correlation with popularity is loudness (0.14). While still weak, it might indicate that songs with higher loudness levels tend to have slightly higher popularity.


```{r ,echo=TRUE,results='markup'}

# Calculate Spearman's rank correlation
cor_matrix_spearman <- cor(spotify_num, use = "complete.obs", method = "spearman")
print("Spearman Correlation Matrix:")
print(cor_matrix_spearman)

# Visualize Spearman correlation
corrplot::corrplot(cor_matrix_spearman, method = "color", addCoef.col = "black", title = "Spearman Correlation", mar = c(0, 0, 1, 0))

# Calculate Kendall's tau correlation
cor_matrix_kendall <- cor(spotify_num, use = "complete.obs", method = "kendall")
print("Kendall's Tau Correlation Matrix:")
print(cor_matrix_kendall)

# Visualize Kendall correlation
corrplot::corrplot(cor_matrix_kendall, method = "color", addCoef.col = "black", title = "Kendall's Tau Correlation", mar = c(0, 0, 1, 0))


```




```{r ,echo=TRUE,results='markup'}
# Build a simple linear model with the strongest correlated variable
lm_model1 <- lm(popularity ~ loudness, data = spotify_num)

# Summarize the model
summary(lm_model1)
```

Strength of Relationship: While loudness is statistically significant, it explains only a small proportion of the variance in popularity, suggesting the need for additional predictors to improve the model.
Practical Implication: Songs with higher loudness tend to be slightly more popular, but other features like danceability, energy, or tempo may have stronger effects.



```{r ,echo=TRUE,results='markup'}
# Build a two-variable model
lm_model2 <- lm(popularity ~ loudness + danceability, data = spotify_num)

# Summarize the model
summary(lm_model2)

# Check Variance Inflation Factor (VIF)
vif(lm_model2)
```
Loudness is a Positive Predictor:

The positive relationship between loudness and popularity remains consistent with the single-variable model, suggesting that louder songs are slightly more likely to be popular.
Danceability is a Negative Predictor:

Surprisingly, danceability has a negative association with popularity. This could indicate that highly danceable songs may not align with broader preferences in the dataset.
Weak Explanatory Power:

The low R-squared value indicates that many other factors influence popularity, and this model alone does not capture the complexity of the relationship.

```{r ,echo=TRUE,results='markup'}
lm_model3 <- lm(popularity ~ loudness + danceability + energy, data = spotify_num)

summary(lm_model3)
vif(lm_model3)

```

```{r ,echo=TRUE,results='markup'}

lm_model4 <- lm(popularity ~ loudness + danceability + energy+tempo, data = spotify_num)

# Summarize the model
summary(lm_model4)

# Check VIF
vif(lm_model4)

```

```{r ,echo=TRUE,results='markup'}

lm_model5 <- lm(popularity ~ loudness + danceability + energy+ tempo+ valence, data = spotify_num)


summary(lm_model5)
vif(lm_model5)

```

```{r ,echo=TRUE,results='markup'}
conf_intervals <- confint(lm_model3)
print(conf_intervals)
```



```{r ,echo=TRUE,results='markup'}
anova_results <- anova(lm_model1, lm_model2, lm_model3, lm_model4, lm_model5)
print(anova_results)

```
```{r ,echo=TRUE,results='markup'}
lm_interaction <- lm(popularity ~ loudness * danceability, data = spotify_num)
summary(lm_interaction)

```
Relationship between loudness and popularity becomes more positive as danceability increases. For highly danceable songs, increasing loudness has a stronger positive impact on popularity.

```{r ,echo=TRUE,results='markup'}
ggplot(spotify_num, aes(x = loudness, y = popularity)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red") +
  labs(title = "Loudness vs Popularity", x = "Loudness", y = "Popularity") +
  theme_minimal()
```