---
title: "DATS6101_Summary_Report_FINAL_PROJECT"
author: "Team 5"
date: "2024-12-13"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    code_folding: hide
---

```{r message=TRUE, include=FALSE, results='asis'}
# Loading the necessary libraries
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(plotly)
library(car)
library(PMCMRplus)
library(here)
```


# Abstract 










# Introduction

#### **Background** 









#### **Objectives**

Analyze the Influence of Musical Attributes:
To examine how musical features like danceability, energy, tempo, valence, and loudness contribute to the popularity of songs, providing insights into key drivers of listener engagement and chart performance.

Evaluate Factors Impacting Top Rankings:
To determine how specific musical features influence a song's likelihood of reaching the top 25 rankings on global charts, identifying attributes that correlate strongly with chart success.

Understand Regional and Cultural Preferences:
To uncover variations in musical preferences across countries and regions, exploring how cultural and geographic factors shape listener behavior and song performance.

Identify Temporal Patterns in Song Popularity:
To investigate trends in the rise and fall of song popularity over time, analyzing seasonal patterns and the lifespan of popular songs to inform optimal release strategies.


#### **Research questions**

The backbone of our study lies in answering the following SMART questions:

Q1: How do musical attributes like danceability, energy, tempo, valence, and loudness contribute to a song’s popularity?

Q2: How do musical features influence a song's likelihood of reaching the top 25 rankings on global charts?

Q3: How do musical preferences vary across countries and regions?

Q4: What temporal patterns exist in rise and fall of song popularity?

*These SMART questions were developed to identify the key factors driving song popularity and to provide actionable insights into the global music landscape. Each question targets specific elements, such as musical attributes, regional preferences, and temporal trends, aiming to uncover patterns that can inform music producers, marketers, and streaming platforms. The insights generated can help optimize song production strategies, tailor content to regional preferences, and identify the ideal timing for music releases to maximize audience engagement and chart performance.*


# Dataset Overview











# Data Preprocessing

```{r, include=T, results='markup',message=TRUE}

```


```{r, include=T, results='markup',message=TRUE}

```


```{r}

```


```{r, include=T, results='markup',message=TRUE}


```



```{r}

```


```{r}


```

# Exploratory Data Analysis













## Research Question 1. 







### Bivariate Analysis and Multivariate Analysis : 



##### Overview of Weather and Lighting Effects:



##### Key Findings:




### **Hypothesis Testing**



### Practical Implications :





## 3. Analysis of Infrastrcture conditions



#### Bivariate Analysis :

#### Multivariate Analysis :

#### **Tests carried out**

#### 1. chi-square test on the effect of road types







# Conclusion

This project successfully identified key drivers of song popularity on Spotify, providing actionable insights into musical and non-musical attributes. By leveraging advanced modeling techniques and robust data analysis, the study highlighted significant regional and temporal patterns, offering valuable guidance for artists, producers, and marketers. Future work could extend this analysis to incorporate additional data sources and explore emerging trends in the music industry.













